---
title: "Equity: who the changes affect"
subtitle: "The case for a different solution for immersion families"
---

```{r}
source("R/site_data.R")
source("R/charts.R")
library(gt)
s44 <- read_data("pps_slide44.csv")
grp <- read_data("equity_groups.csv")
chg <- read_data("equity_changes.csv")
home <- read_data("equity_home_distance.csv")
homes <- read_data("equity_home_distance_summary.csv")
comp <- read_data("equity_school_composition.csv")
dem <- read_data("equity_demand_moves.csv")
mlm <- read_data("equity_ml_moving.csv")
opt <- read_data("equity_optout.csv")
acc <- read_data("access_summary.csv") |> filter(group == "Spanish (ages 5-17)")
acc_ci <- read_data("access_robustness_ci.csv")
g <- function(sc, ty, col) grp[[col]][grp$scenario == sc & grp$type == ty]
cp <- function(sc, grp_name, col) comp[[col]][comp$scenario == sc & comp$group == grp_name]
DLI <- "Immersion program moves"; CLO <- "School closures (non-immersion students)"; ALL <- "All PPS students"
SEND <- "Schools losing an immersion program"; RECV <- "Schools receiving an immersion program"
CLOS <- "Schools closing (no immersion program)"; DIST <- "All PPS schools"
h <- function(sc, ty, col) homes[[col]][homes$scenario == sc & homes$type == ty]
hck <- read_data("equity_home_check_summary.csv")
hc <- function(sc, col) hck[[col]][hck$scenario == sc]
d_a <- dem |> filter(scenario == "a")
dm <- function(school, col) d_a[[col]][d_a$from_school == school][1]
opt_tot <- opt |> filter(scenario == "a") |> group_by(assumption) |>
  summarise(rate = first(rate), lost = sum(students_lost), k = sum(k_lost_per_year, na.rm = TRUE))
ot <- function(a, col) opt_tot[[col]][opt_tot$assumption == a]
prec <- "Bridger to Lent precedent"
displaced_a <- g("a", DLI, "students") + g("a", CLO, "students")
fol <- read_data("equity_pps_follow.csv")
fol1 <- fol |> filter(year == "2027-28")
ses <- read_data("equity_se_seats.csv"); sx <- function(k, col) ses[[col]][ses$key == k]
sctx <- read_data("equity_site_context.csv"); sc <- function(k, col) sctx[[col]][sctx$key == k]
rig_own <- sites |> filter(school_key == "rigler", language == "Spanish", band == "K-5") |> select(scenario, own_area)
ge_r <- read_data("guide_enrollment.csv") |> filter(school == "Rigler", year %in% c("2027-28", "2031-32"))
gr <- function(sc, yr) ge_r$enrollment[ge_r$scenario == sc & ge_r$year == yr]
gs_r <- read_data("guide_school.csv") |> filter(school == "Rigler")
gsr <- function(opt, col) gs_r[[col]][gs_r$option == opt]
capt <- read_data("equity_capture.csv")
cv <- function(gr, yr, col = "capture") capt[[col]][capt$group == gr & capt$year == yr]
SE <- "Southeast (excluding Lent)"; REST <- "Rest of PPS"
fol_range <- sprintf("%.0f–%.0f%%", 100 * min(fol1$followed), 100 * max(fol1$followed))
```

::: {.callout-important appearance="simple"}
## The argument in brief

- **The plan leans on moving immersion programs.** Scenario A moves `r fmt_n(g("a", DLI, "students"))` immersion students out of their schools. That is `r fmt_pct(g("a", DLI, "students") / displaced_a)` of the students directly displaced by a closure or program move ([§1](#sec-relies)).
- **The programs being moved are not short of families.** Atkinson draws `r fmt_n(dm("Atkinson", "applications"))` kindergarten applications a year for about `r fmt_n(dm("Atkinson", "offered"))` seats, and Beach `r fmt_n(dm("Beach", "applications"))` for `r fmt_n(dm("Beach", "offered"))` ([§2](#sec-demand)).
- **The burden falls on schools serving underserved communities.** Schools losing an immersion program are `r fmt_pct(cp("a", SEND, "pct_hist_underserved"))` historically underserved and `r fmt_pct(cp("a", SEND, "pct_direct_cert"))` low-income. The schools closing outright are `r fmt_pct(cp("a", CLOS, "pct_white"))` white, with `r fmt_pct(cp("a", CLOS, "pct_hist_underserved"))` underserved. Students in the moving programs are `r round(g("a", DLI, "pct_spanish_home") / g("a", CLO, "pct_spanish_home"))` times as likely to speak Spanish at home as students at closing schools ([§3](#sec-who)).
- **The moves are longer.** Their schools move farther (`r sprintf("%.1f", g("a", DLI, "school_miles"))` vs `r sprintf("%.1f", g("a", CLO, "school_miles"))` miles), and the share of school-age Spanish speakers within a mile of a K-5 Spanish site falls from `r fmt_pct(acc$within_1mi[acc$scenario == "sq"])` to `r fmt_pct(acc$within_1mi[acc$scenario == "a"])` ([§4](#sec-burden)).
- **Moving a program loses the families who would have joined it.** After PPS moved Bridger's Spanish program to Lent in 2023, kindergarten immersion enrollment fell 31%, against 14% district-wide. The 2023 plan had projected Lent at 527 students; it has `r fmt_n(sum(by_grade$enrollment[by_grade$key == "lent" & by_grade$year == "2025-26"]))`. PPS's own school projections now expect only `r fol_range` of students in a moved program to follow it; the rest stay at neighborhood schools. By PPS's own numbers, the moves shrink immersion ([§5](#sec-optout)).
- **The last round of changes moved neighborhood enrollment too.** Since the 2023 Southeast changes, the share of Southeast elementary children attending their own neighborhood school fell from `r fmt_pct(cv(SE, "2022-23"))` to `r fmt_pct(cv(SE, "2025-26"))`, while it rose elsewhere in PPS. Public data can't separate how much of that drop was built into the changes and how much was families' choice ([§6](#sec-capture)).
- **PPS's equity check can't see any of this.** It reports one move rate per group. Even so, multilingual learners are already the most-moved group ([§7](#sec-pps)).
- **Southeast Spanish immersion is undersized for its demand.** Richmond alone enrolls `r sx("richmond", "k_2025")` Japanese kindergartners a year, more than Southeast's two Spanish programs combined (`r sx("lent", "k_2025") + sx("atkinson", "k_2025")`), while Atkinson turns away about two applicants for every one it admits. Lent's problem is that it no longer serves its own neighborhood, not a shortage of families who want Spanish, and the plan would do the same to Rigler ([§8](#sec-southeast)).
:::

## 1. The plan leans on moving immersion programs {#sec-relies}

PPS's scenario memo names "underenrolled dual language immersion classrooms" among the problems the modeling responds to. It calls immersion consolidation a "primary driver" of the changes in the Roosevelt/Jefferson region (`r SRC$memo`). Both scenarios cut elementary Spanish immersion sites from nine to four and move two Mandarin programs. Scenario A also moves Vietnamese immersion.

In Scenario A, `r fmt_n(g("a", DLI, "students"))` students are in immersion programs that move. Another `r fmt_n(g("a", CLO, "students"))` attend schools that close. So immersion students are **`r fmt_pct(g("a", DLI, "students") / displaced_a)` of everyone directly displaced**, while immersion students are `r fmt_pct(cp("a", DIST, "pct_immersion"))` of PPS enrollment. Boundary changes, which move many more students, are not counted in either group. Their makeup can't be measured from public data.

## 2. "Underenrolled" describes seats, not demand {#sec-demand}

```{r}
#| label: tbl-demand
d_a |>
  arrange(desc(applications)) |>
  transmute(move = paste(from_school, "→", to_school), language, applications, offered, waitlisted, ratio = applications / offered) |>
  gt() |>
  cols_label(move = "Program move", language = "Language", applications = "Applications", offered = "Offers",
             waitlisted = "Waitlisted", ratio = "Applications per offer") |>
  fmt_number(c(applications, offered, waitlisted), decimals = 0) |> fmt_number(ratio, decimals = 1) |>
  tab_spanner("Kindergarten lottery, average per year", c(applications, offered, waitlisted, ratio)) |>
  tab_style(cell_text(weight = "bold"), cells_body(rows = ratio >= 3)) |>
  tab_header(title = "Demand at the programs being moved (Scenario A)") |>
  tab_source_note("Average of the 2022-23 to 2026-27 lotteries (fewer years where a program wasn't in the lottery). Counts are applications, not children: a family can apply to several programs.") |>
  tab_source_note(md(cite("lottery", "scenarios")))
```

The two Spanish programs with the most applicants in the district, **Atkinson and Beach**, each get four to five applications for every kindergarten seat, and dozens of families are waitlisted every year. Their classrooms are small because PPS offers few seats, not because families aren't asking for them. Other moves are different. James John and Sitton draw little more than one application per seat, and Clark (whose applicants are often placed at programs they ranked higher) has had no waitlist. The case for consolidating those programs is stronger. A plan that treated these programs differently would take less from immersion families than one that moves them all. More on demand, including why Lent's own lottery numbers understate its draw, is on the [Demand & scorecard](scorecard.qmd) page.

## 3. Who the burden falls on {#sec-who}

```{r}
#| label: tbl-composition
comp |> filter(scenario == "a") |>
  mutate(group = factor(group, c(SEND, CLOS, RECV, DIST))) |> arrange(group) |>
  select(group, enrollment, pct_hist_underserved, pct_direct_cert, pct_latino, pct_white, pct_ell) |>
  gt() |>
  cols_label(group = "", enrollment = "Students", pct_hist_underserved = "Historically underserved",
             pct_direct_cert = "Low-income (direct certification)", pct_latino = "Latino", pct_white = "White", pct_ell = "English learners") |>
  fmt_number(enrollment, decimals = 0) |> fmt_percent(starts_with("pct"), decimals = 0) |>
  tab_style(cell_text(weight = "bold"), cells_body(rows = group %in% c(SEND, CLOS))) |>
  tab_header(title = "Who attends the schools involved (Scenario A)", subtitle = "Whole-school composition, October 2025, weighted by enrollment") |>
  tab_source_note("PPS defines historically underserved students as Latino, African American, Native American, Pacific Islander and multi-racial (other ancestries). Direct certification: students the state has identified as eligible for free meals. Each school is counted once: Beach, James John and Rose City Park, which close and lose an immersion program, are counted with schools losing an immersion program. Clark is not printed in PPS's report and is left out. These are whole-school figures; PPS does not publish race or income by program.") |>
  tab_source_note(md(cite("demographics", "scenarios", "closures")))
```

The schools whose immersion programs move are **`r fmt_pct(cp("a", SEND, "pct_hist_underserved"))` historically underserved and `r fmt_pct(cp("a", SEND, "pct_direct_cert"))` low-income**. The district averages are `r fmt_pct(cp("a", DIST, "pct_hist_underserved"))` and `r fmt_pct(cp("a", DIST, "pct_direct_cert"))`. The schools that close outright are `r fmt_pct(cp("a", CLOS, "pct_white"))` white, `r fmt_pct(cp("a", CLOS, "pct_hist_underserved"))` underserved and `r fmt_pct(cp("a", CLOS, "pct_direct_cert"))` low-income. Scenario B shows the same pattern (`r fmt_pct(cp("b", SEND, "pct_hist_underserved"))` vs `r fmt_pct(cp("b", CLOS, "pct_hist_underserved"))` underserved).

These race and income figures describe **whole schools**, including students in their neighborhood programs, who don't move. PPS doesn't publish race or income for immersion programs alone, so they don't show that the students who move are half underserved. The moving students themselves can be measured by home language and English-learner status:

```{r}
#| label: fig-who
#| fig-height: 4.6
#| fig-cap: "Share of students whose home language is Spanish, or any language other than English: students in immersion programs that move, students at schools that close (excluding their immersion programs), and all PPS students. October 2025 enrollment."
d <- grp |> mutate(type = factor(type, c(DLI, CLO, ALL), c("Immersion program moves", "School closures", "All PPS students")),
                   scen = paste0("Scenario ", toupper(scenario))) |>
  filter(!(type == "All PPS students" & scenario == "b")) |>
  mutate(scen = if_else(type == "All PPS students", "District", scen), label = paste0(type, if_else(scen == "District", "", paste0(", ", scen))))
d$label <- factor(wrap_label(d$label), rev(unique(wrap_label(d$label)[order(d$type, d$scen)])))
tip <- function(x, what) sprintf("<b>%s</b><br>%s%% %s<br>%s students", d$label, round(100 * x), what, fmt_n(d$students))
p <- plot_ly(d, y = ~label) |>
  add_bars(x = ~pct_spanish_home, name = "Spanish at home", marker = list(color = PAL$light[["blue"]], line = list(color = "white", width = 2)),
           text = tip(d$pct_spanish_home, "speak Spanish at home"), hoverinfo = "text", textposition = "none") |>
  add_bars(x = ~pct_non_english_home, name = "Any non-English language at home", marker = list(color = PAL$light[["orange"]], line = list(color = "white", width = 2)),
           text = tip(d$pct_non_english_home, "speak a language other than English at home"), hoverinfo = "text", textposition = "none")
viz_layout(p, xtitle = "Share of students") |>
  layout(barmode = "group", bargap = 0.3, xaxis = list(tickformat = ".0%", rangemode = "tozero"))
```

::: {.source}
`r cite("by_lang", "by_grade", "scenarios", "closures")`
:::

In Scenario A, **`r fmt_pct(g("a", DLI, "pct_spanish_home"))` of students in the moving programs speak Spanish at home** and `r fmt_pct(g("a", DLI, "pct_non_english_home"))` speak a language other than English. At the closing schools the figures are `r fmt_pct(g("a", CLO, "pct_spanish_home"))` and `r fmt_pct(g("a", CLO, "pct_non_english_home"))`. `r fmt_pct(mlm$pct_ml[mlm$scenario == "a"])` of students in the moving programs are multilingual learners still receiving English language services. Of the Spanish-speaking students directly displaced in Scenario A, **`r fmt_pct(g("a", DLI, "spanish_home") / (g("a", DLI, "spanish_home") + g("a", CLO, "spanish_home")))` are immersion students**.

These are the students two-way immersion depends on. The model needs native speakers of the program language in every classroom, alongside English speakers.

## 4. Immersion programs move farther from families {#sec-burden}

**How far the school moves (measured).** Students at closing schools mostly move to the next school over, on average `r sprintf("%.1f", g("a", CLO, "school_miles"))` miles. Immersion programs move farther: `r sprintf("%.1f", g("a", DLI, "school_miles"))` miles on average. Atkinson, Beach, James John and Sitton each move 1.6–2.5 miles; Scott → Rigler, at 0.7 miles, is the exception.

```{r}
#| label: tbl-school-miles
grp |> filter(type != ALL) |>
  mutate(type = if_else(type == DLI, "Immersion program moves", "School closures"), scenario = paste("Scenario", toupper(scenario))) |>
  select(type, scenario, school_miles) |>
  pivot_wider(names_from = scenario, values_from = school_miles) |>
  gt() |>
  cols_label(type = "") |>
  fmt_number(-type, decimals = 1) |>
  tab_header(title = "How far the school moves, miles (straight line)", subtitle = "Average over affected students") |>
  tab_source_note("Distance from the current school to the new school, averaged over each change's enrolled students (October 2025). For closures split among several receiving schools, weighted by the share of the closing school's attendance area each receives.") |>
  tab_source_note(md(cite("scenarios", "by_grade", "by_lang")))
```

**Access (estimate).** Today `r fmt_pct(acc$within_1mi[acc$scenario == "sq"])` of school-age Spanish speakers live within a mile of a K-5 Spanish immersion site. Under either scenario, `r fmt_pct(acc$within_1mi[acc$scenario == "a"])` would. Allowing for Census survey error, the drop is `r round(100 * acc_ci$lo[acc_ci$measure == "drop_points"])`–`r round(100 * acc_ci$hi[acc_ci$measure == "drop_points"])` percentage points.

**Distance from home (model estimate).** PPS doesn't publish where students live. Children are placed where the 2020 Census counted them. For immersion programs, families are assumed to live in the area nearest their current site. On that basis, home-to-school distance grows by about `r sprintf("%.1f", h("a", DLI, "change"))` mile for immersion families and `r sprintf("%.1f", h("a", CLO, "change"))` mile for families at closing schools. PPS's report of where each program's students live gives a check: `r fmt_pct(hc("a", "own_share"))` of students in the moving programs live in their school's own attendance area, and the rest are reported only by high-school area. Placing them there gives an increase of `r sprintf("%.1f", hc("a", "change_mi"))` to `r sprintf("%.1f", hc("a", "change_mi_nearer"))` miles, close to the model, although the model overstates how far families live from school today (`r sprintf("%.1f", hc("a", "model_before_mi"))` miles against `r sprintf("%.1f", hc("a", "before_mi"))`). The [details and sensitivity checks](#sec-distance-detail) are at the end of this page.

## 5. Moving a program loses the families who would have joined it {#sec-optout}

When a program moves farther away, some families choose their neighborhood school instead. PPS staff described this at a board committee meeting in September: "when we move, for example, a program … we did not see students necessarily leave the district, but they may not have moved with the program" (`r SRC$tle0908`). PPS's equity audit of the 2023 Southeast changes found that students in moved programs who stayed in PPS generally followed their program (`r SRC$audit`, pp. 11, 48–49). Its description of who left the moved programs is less clear. It says leavers "more often identified as white" (p. 11). The audit reports Latino/a/x students as a separate group alongside White, Asian and Black students, so this most likely means non-Latino white students, but the report does not say how students who are both white and Latino were counted. On home language the audit contradicts itself: leavers "more often speak a language other than English at home" (p. 11), but "families who left were disproportionately English-speaking" (p. 49). Because enrolled students mostly followed their program, the larger loss is likely among families who haven't joined yet: kindergarten families deciding whether to choose a program that is now farther away.

The clearest test is the last time PPS consolidated Spanish immersion. In 2023-24, **Bridger's Spanish program moved to Lent** (`r SRC$summary10`). Before the move the two schools had about 55 immersion students per grade. Lent has since enrolled 32 and 43 immersion kindergartners. That is **31% fewer**, while kindergarten enrollment district-wide fell 14% on the same basis. That works out to about **one in five families lost to the program beyond the district trend** ([details](scorecard.qmd#a-precedent-bridgers-spanish-program-moved-to-lent-in-2023)). The audit also found that the Southeast projections overestimated enrollment at 12 of 20 schools. Of the 7 schools still well short of target, 3 had a program relocation or grade change (`r SRC$audit`, p. 6).

The 2023 plan itself expected Lent to grow, not shrink. PPS's Phase 2 impact analysis, as summarized in the audit's school-by-school appendix, projected that converting Lent to whole-school Spanish immersion "by adding two strands of Spanish immersion from Bridger" would take it "from 277 students in 2021–22 to an estimated 527 students in 2025–26", with building use rising "from 43.8% to 87.5%" (`r SRC$audit_schools`, p. 34). In October 2025 Lent enrolled `r fmt_n(sum(by_grade$enrollment[by_grade$key == "lent" & by_grade$year == "2025-26"]))` students, about half that projection. (The appendix doesn't say whether the 527 includes pre-kindergarten; even with PPS's pre-kindergarten counted, Lent is far short.) Of the 265 students at Lent the spring before the conversion, neighborhood and immersion alike, 49 (18.5%) had left PPS within two years (p. 45).

PSU's enrollment forecasts did not anticipate the shortfall either. The forecast PSU prepared for PPS in June 2023 built the Bridger → Lent move in (its notes read "Bridger Spanish DLI (grades K-5) moved to Lent ES"). It kept almost all of the combined program's 319 students: 314 for Lent in 2023-24. Lent enrolled 273. For three years running the forecast for Lent was `r fcl <- read_data("equity_forecast_check.csv"); a <- filter(fcl, target == "Lent", vintage == "2023-24"); d <- filter(fcl, target == "District", vintage == "2023-24"); sprintf("%.0f–%.0f%% too high, while its district-wide forecast was within %.0f%%", 100 * min(a$pct_error), 100 * max(a$pct_error), 100 * max(abs(d$pct_error)))`.

```{r}
#| label: tbl-forecast
read_data("equity_forecast_check.csv") |>
  filter(vintage == "2023-24" | target == "Lent") |>
  mutate(row = case_when(target == "Lent" & vintage == "2023-24" ~ "Lent: forecast made with the move built in (June 2023)",
                         target == "Lent" ~ "Bridger + Lent Spanish: forecast made before the move (June 2022)",
                         TRUE ~ "District total: same June 2023 forecast")) |>
  select(row, year, forecast, actual, pct_error) |>
  pivot_wider(names_from = year, values_from = c(forecast, actual, pct_error), names_vary = "slowest") |>
  gt() |>
  cols_label(.list = c(list(row = ""), setNames(as.list(rep(c("Forecast", "Actual", "Error"), 3)),
                                        paste0(rep(c("forecast_", "actual_", "pct_error_"), 3), rep(c("2023-24", "2024-25", "2025-26"), each = 3))))) |>
  tab_spanner("2023-24", ends_with("2023-24")) |> tab_spanner("2024-25", ends_with("2024-25")) |> tab_spanner("2025-26", ends_with("2025-26")) |>
  fmt_number(c(starts_with("forecast"), starts_with("actual")), decimals = 0) |>
  fmt_percent(starts_with("pct_error"), decimals = 0, force_sign = TRUE) |>
  cols_width(row ~ px(240)) |>
  tab_header(title = "Lent's enrollment fell well short of PPS's forecast") |>
  tab_source_note("Forecasts and October actual enrollment (kindergarten through grade 12) from the PSU Population Research Center forecasts for PPS. Error = (forecast − actual) ÷ actual. Other schools also miss their forecasts, in both directions; Lent's miss is unusually large and lasting for a school whose program was moved.") |>
  tab_source_note(md(cite("psu_premove")))
```

This time PPS's projections do expect some families not to follow. PPS's per-school guides for the scenarios project each school's enrollment with no changes and under the options (`r SRC$guides`). Their "no changes" figures match PSU's 2026-27 forecast exactly, so the arrival of a moved program can be read against PSU's forecast of that program. In the three moves where nothing else changes at the receiving school, PPS projects that `r fol_range` of the program's students arrive. Where the guides show the sending school, the rest stay there as neighborhood students:

```{r}
#| label: tbl-pps-follow
fol |> filter(year %in% c("2027-28", "2031-32")) |>
  mutate(move = paste0(move, " (", language, ", Scenario ", toupper(scenario), ")")) |>
  select(move, year, program, joined, followed, stayed_n, stayed) |>
  gt(groupname_col = "move") |>
  cols_label(year = "Year", program = "Program size (PSU forecast)", joined = "Arrive at new school",
             followed = "Share arriving", stayed_n = "Stay at old school", stayed = "Share staying") |>
  fmt_number(c(program, joined, stayed_n), decimals = 0) |> fmt_percent(c(followed, stayed), decimals = 0) |>
  sub_missing(missing_text = "–") |>
  tab_header(title = "PPS's projections have about one in five students not following a moved program") |>
  tab_source_note("Arrive = receiving school's projected enrollment under the options minus its projection with no changes. Stay = sending school's projection under the options minus PSU's forecast of its other programs (shown only where the sending school has no other change: Scott in Scenario B, Clark). Program size = PSU's 2026-27 forecast of the moving program in that year. Other moves (to César Chávez, Sitton and Vestal) come with boundary or grade changes and can't be separated. The guides don't state a follow rate; these shares are implied by their numbers.") |>
  tab_source_note(md(cite("guides", "psu_premove")))
```

That is close to the Bridger → Lent precedent, and far from the roughly 98% PSU assumed for that move in 2023. It also means the loss is part of the plan: PPS expects the consolidated programs to be smaller than the programs they replace.

```{r}
#| label: tbl-optout
opt |> filter(scenario == "a") |>
  group_by(assumption) |>
  summarise(rate = first(rate), students = sum(students), lost = sum(students_lost), k = sum(k_lost_per_year, na.rm = TRUE), .groups = "drop") |>
  arrange(rate) |>
  gt() |>
  cols_label(assumption = "If this share of families doesn't follow the program", rate = "Share",
             students = "Students in moving programs", lost = "Students not following", k = "Kindergartners not joining, per year") |>
  fmt_percent(rate, decimals = 0) |> fmt_number(c(students, lost, k), decimals = 0) |>
  tab_header(title = "Families who might choose a neighborhood school instead (Scenario A)", subtitle = "Estimate: current students and yearly kindergarten entry in the moving programs, under three assumptions") |>
  tab_source_note("Students: October 2025 enrollment in each moving program, in the grades that move (program moves only). Kindergartners: October 2025 kindergarten immersion enrollment at each sending K-5 program. The 7% row uses a figure a public commenter attributed to PPS at the October 6, 2026 board meeting (PPS's estimate of families who would leave the district); we could not find it in PPS documents. The precedent row applies the excess loss measured after the Bridger to Lent move.") |>
  tab_source_note(md(cite("by_grade", "scenarios", "profiles", "class_size", "summary10")))
```

If Scenario A loses families at the Bridger → Lent rate, about **`r fmt_n(ot(prec, "k"))` kindergartners a year** who would have joined these programs go elsewhere. The figure at 7% is `r fmt_n(ot("7% (figure cited in public comment)", "k"))`. Over six years of kindergarten classes that adds up to roughly `r fmt_n(6 * ot(prec, "k"))` children. Most would be expected to enroll at their neighborhood school instead. PPS's guides place these students at neighborhood schools and flag whether each school meets its size and space goals, but they give no capacity figures. PPS publishes each building's current capacity separately, in its School Profiles dashboard ([§8](#sec-southeast)). Either way, the students lost would come from programs at schools that are about half historically underserved ([§3](#sec-who)).

### What staff and families at Lent and Bridger described

The audit's interviews and surveys describe what the 2023 move was like for the schools involved. The appendix identifies speakers only as participants or staff at each school.

- **Families who pushed for the move didn't all stay.** A Lent participant: "A lot of the parents who pushed from Bridger for their kids to come to the school didn't end up staying…they felt like it was out of their way [and that] their students didn't feel supported" (`r SRC$audit_schools`, p. 39).
- **Promises made to families weren't kept.** A Lent participant: "the fourth grade parents were really upset… they were promised it would be a three strand school…and the parents [started] bringing the letters that they were promised" (p. 41). Families in the main report described transition staffing or funding that was "reduced, delayed, or pulled back because of budget constraints" (`r SRC$audit`, p. 83).
- **The transition was hard on staff and the neighborhood.** "Those years were really terrible, those first few years…There were a lot of tears during the transition" (appendix, p. 36). After 2022-23, 63% of Lent staff left the school, against a 32% average across the affected schools (p. 39). Staff described hostility from neighborhood families who lost access: "We have been cussed out. We have been yelled at…by parents who moved into the neighborhood that do not want their students to speak Spanish, and that feel that it's unfair that their students can't go to their neighborhood school" (p. 41), and the loss of a walkable school: "We had a walking school bus every day…We don't have that anymore, because our entire population gets on a bus to go home at the end of the day" (p. 43).
- **Families who moved with the program paid in travel.** "Families who moved from Bridger to Lent described losing walkable access to school, longer commutes, early bus times, transportation uncertainty, and a loss of neighborhood connection" (main report, p. 83).
- **Bridger changed too.** Participants "experienced the Bridger student body as becoming more white and more affluent, while also losing the additional staffing and Title I support that had previously accompanied a higher-poverty enrollment" (appendix, p. 17).
- **Who the consolidation served was questioned.** Marysville participants "noted that the DLI strand was concentrated in a school farther from many Latino/a/x families, while Atkinson, located in a whiter and more affluent neighborhood, retained a single-strand DLI program", and observed "that some immigrant families may have been hesitant to enroll their children in full Spanish immersion in the broader political climate" (p. 61).

Parents summed up what the move needed: "Whole school DLI is great. It needs to be funded, and it needs to be prioritized" (main report, p. 84).

## 6. Where neighborhood children go {#sec-capture}

The scenario projections place most students at the schools the new boundaries and program sites assign them. PPS tracks how often that happens: the capture rate is the share of children living in a school's attendance area who attend that school. District-wide it has held near `r fmt_pct(cv("All PPS, K-12", "2025-26"))` for years, so about a third of PPS students already attend somewhere other than their neighborhood school.

```{r}
#| label: fig-capture
#| fig-height: 4.2
#| fig-cap: "Neighborhood capture rate: share of K-12 students living in each elementary attendance area who attend that area's school, October counts. Southeast = the 16 elementary areas around the 2023 Southeast changes, excluding Lent. Hover for counts."
cg <- capt |> filter(group %in% c(SE, REST, "Lent")) |>
  mutate(group = factor(group, c(SE, REST, "Lent"), c("Southeast elementary areas (excl. Lent)", "Rest of PPS elementary areas", "Lent area")))
colr <- c(PAL$light[["blue"]], PAL$muted, PAL$light[["orange"]])
p <- plot_ly()
for (i in seq_along(levels(cg$group))) {
  d <- filter(cg, group == levels(cg$group)[i])
  p <- add_trace(p, data = d, x = ~year, y = ~capture, type = "scatter", mode = "lines+markers", name = levels(cg$group)[i],
                 line = list(color = colr[i], width = 2), marker = list(color = colr[i], size = 8),
                 text = ~sprintf("<b>%s</b> %s<br>%s of %s children attend their own neighborhood school (%s)", group, year,
                                 fmt_n(own_neighborhood_school), fmt_n(total), fmt_pct(capture)), hoverinfo = "text")
}
p <- add_segments(p, x = "2022-23", xend = "2022-23", y = 0, yend = 0.9, line = list(color = PAL$grid, width = 1, dash = "dot"), showlegend = FALSE, hoverinfo = "skip")
viz_layout(p, ytitle = "Share attending own neighborhood school") |>
  layout(xaxis = list(type = "category"), yaxis = list(tickformat = ".0%", range = c(0, 0.92)),
         annotations = list(list(x = "2022-23", y = 0.9, text = "2023 Southeast changes<br>take effect next year", align = "left", showarrow = FALSE, xanchor = "left", xshift = 4, font = list(size = 11, color = PAL$ink2))))
```

::: {.source}
`r cite("capture")`
:::

After the 2023 Southeast changes, capture in the Southeast elementary areas fell from `r fmt_pct(cv(SE, "2022-23"))` to `r fmt_pct(cv(SE, "2023-24"))` and has stayed there (`r fmt_pct(cv(SE, "2025-26"))` in 2025-26), while the rest of PPS rose from `r fmt_pct(cv(REST, "2022-23"))` to `r fmt_pct(cv(REST, "2025-26"))`. In the Southeast, the share attending PPS alternative and focus schools rose from `r fmt_pct(cv(SE, "2022-23", "alternative"))` to `r fmt_pct(cv(SE, "2025-26", "alternative"))`. Some of this drop was built into the changes and doesn't reflect families' choices. Children allowed to finish at their old school after a boundary moved count as attending "another neighborhood school", and Creative Science, a focus school, moved into the Bridger building, which put an alternative school inside the Southeast. PPS's reports don't break the drop down, so the share due to families' choice can't be measured. If it were mostly students finishing at their old schools, it should be fading as they move up; three years on, Southeast capture has recovered by only `r sprintf("%.1f", 100 * (cv(SE, "2025-26") - cv(SE, "2023-24")))` percentage points.

The Lent area shows what a whole-school immersion site does to its neighborhood. Capture there fell from `r fmt_pct(cv("Lent", "2022-23"))` to `r fmt_pct(cv("Lent", "2023-24"))` when Lent converted. That drop is mostly by design, not a sign of families leaving: Lent's non-immersion neighborhood children were assigned to Marysville, and PPS's report counts them as attending "another neighborhood school". Today `r fmt_n(cv("Lent", "2025-26", "own_neighborhood_school"))` of Lent's `r fmt_n(sum(by_grade$enrollment[by_grade$key == "lent" & by_grade$year == "2025-26"]))` students live in its area, so most of its families travel from elsewhere. In both scenarios three of the four K-5 Spanish sites (Lent, Rigler and César Chávez) would be whole-school immersion sites of this kind.

## 7. Why PPS's equity check misses it {#sec-pps}

PPS's equity check is one slide: the share of kindergarten–8th grade students in each group who would change schools.

```{r}
#| label: tbl-slide44
s44 |>
  gt() |>
  cols_label(group = "Student group", a_number = "Students", a_pct = "Share", b_number = "Students", b_pct = "Share") |>
  tab_spanner("Scenario A", c(a_number, a_pct)) |> tab_spanner("Scenario B", c(b_number, b_pct)) |>
  fmt_number(c(a_number, b_number), decimals = 0) |> fmt_percent(c(a_pct, b_pct), decimals = 0) |>
  tab_style(cell_text(weight = "bold"), cells_body(rows = group == "Multilingual Learners")) |>
  tab_source_note("As printed by PPS. The slide gives no year and does not define \"move\" (for example, whether a student whose immersion program relocates counts). Groups overlap.") |>
  tab_source_note(md(cite("slide44")))
```

Even this table shows **multilingual learners as the most-moved group** (31% in A, 28% in B, against 27% and 24% overall). It still understates the imbalance:

- It averages each group over all kindergarten–8th grade students, most of whom are moved by boundary changes. That hides what this page shows: closures fall mostly on whiter, higher-income schools, while program moves fall on schools that are about half underserved, and on students far more likely to speak Spanish at home.
- A move to the next school over and a program relocated across town count the same.
- It has no row for immersion students, Latino students or home language. It does not measure distance, and it does not account for families who stop following a program.

## 8. Southeast Spanish immersion is undersized for its demand {#sec-southeast}

In Southeast Portland, PPS offers more kindergarten seats in Japanese than in Spanish. Richmond, a whole-school Japanese immersion school, enrolled `r sx("richmond", "k_2025")` kindergartners in October 2025 (`r fmt_n(sx("richmond", "school_total"))` students in all). Southeast's two Spanish programs, Lent and Atkinson, enrolled `r sx("lent", "k_2025") + sx("atkinson", "k_2025")` between them. Le Monde, a French immersion charter school, enrolled `r sx("lemonde", "k_2025")`, more than any Spanish program in the district except Rigler. District-wide, Spanish still has the most immersion seats; the gap is in the Southeast, home to the district's most oversubscribed Spanish program.

The only other Spanish immersion option nearby is private. The International School of Portland, in South Waterfront across the river from inner Southeast, runs Spanish, Chinese and Japanese immersion tracks and charges tuition. It enrolled 376 students from preschool through 5th grade in 2023-24, including 72 kindergartners, but doesn't publish enrollment by language (`r SRC$isp`).

```{r}
#| label: tbl-se-seats
ses |>
  select(program, k_2025, school_total, applications, offered, waitlisted) |>
  mutate(ratio = applications / offered) |>
  gt() |>
  cols_label(program = "", k_2025 = "Kindergartners, Oct 2025", school_total = "Program students, Oct 2025",
             applications = "Applications", offered = "Offers", waitlisted = "Waitlisted", ratio = "Applications per offer") |>
  tab_spanner("Kindergarten lottery, average per year", c(applications, offered, waitlisted, ratio)) |>
  fmt_number(c(k_2025, school_total, applications, offered, waitlisted), decimals = 0) |> fmt_number(ratio, decimals = 1) |>
  sub_missing(missing_text = "–") |>
  tab_style(cell_text(weight = "bold"), cells_body(rows = program == "Atkinson (Spanish)")) |>
  tab_header(title = "Southeast immersion: kindergarten seats and demand") |>
  tab_source_note("Lottery: 2022-23 to 2026-27 average. Le Monde is a charter school and runs its own admissions, so it isn't in the PPS lottery reports. Lent's lottery counts only families from outside its neighborhood (see the Demand & scorecard page). Applications are not children: a family can apply to several programs.") |>
  tab_source_note(md(cite("by_grade", "lottery")))
```

Richmond admits most of the families who apply; Atkinson admits about one in five. Atkinson's waitlist alone, about `r fmt_n(sx("atkinson", "waitlisted"))` applications a year, is larger than its kindergarten class. Because families can apply to several programs, this doesn't count children exactly, but it suggests a second Atkinson class each year could likely be filled from the waitlist alone. PPS's plan moves this program to Lent instead, and its school guide gives the reason: "Moving Atkinson DLI to Lent brings Lent's student numbers above steady target levels (400+ students)" (`r SRC$guides`).

**Lent's problem is its neighborhood, not its distance from the center.** Lent sits about `r sprintf("%.1f", sc("lent", "edge_mi"))` miles from the district's edge, and families describe students crossing "205, 92nd and 82nd Ave" to reach Marysville, its old neighborhood program's new home (`r SRC$audit_schools`, p. 46). But Rigler is also about `r sprintf("%.1f", sc("rigler", "edge_mi"))` miles from the edge, and it works as a whole-school Spanish immersion school that serves its own neighborhood: `r sc("rigler", "own_area")` of its `r sc("rigler", "students")` students live in its attendance area. At Lent, after its neighborhood program was moved out, `r sc("lent", "own_area")` of `r sc("lent", "students")` do. Lent's surroundings aren't short of Spanish speakers either: about `r fmt_n(sc("lent", "spanish_517_within_1_5mi"))` school-age Spanish speakers live within 1.5 miles of it, against `r fmt_n(sc("atkinson", "spanish_517_within_1_5mi"))` near Atkinson. What Lent lacks is the English-speaking families who fill Atkinson's waitlist, and they live closer to Atkinson.

```{r}
#| label: tbl-sites-context
sctx |> filter(key %in% c("lent", "rigler", "atkinson")) |>
  mutate(school = factor(school, c("Rigler", "Lent", "Atkinson")), own = own_area / students) |> arrange(school) |>
  select(school, edge_mi, students, own_area, own, spanish_517_within_1_5mi) |>
  gt() |>
  cols_label(school = "", edge_mi = "Miles to district edge", students = "Spanish immersion students", own_area = "Live in school's own area",
             own = "Share", spanish_517_within_1_5mi = "School-age Spanish speakers within 1.5 mi") |>
  fmt_number(edge_mi, decimals = 1) |> fmt_number(c(students, own_area, spanish_517_within_1_5mi), decimals = 0) |> fmt_percent(own, decimals = 0) |>
  tab_header(title = "Two whole-school Spanish sites at the district's edge", subtitle = "October 2025; Atkinson for comparison") |>
  tab_source_note("Distances are straight lines from the school's map location to the PPS boundary. School-age Spanish speakers are ACS estimates spread over census blocks (see methods).") |>
  tab_source_note(md(cite("by_residence", "scenarios", "acs", "blocks")))
```

**The plan would make Rigler more like Lent.** Rigler's guide (p. 3) says it "would remain open as a focused regional Spanish DLI K-5 school" and that "Rigler's non-DLI neighborhood boundary would shift to Scott K-5" (`r SRC$guides`). On the scenario maps Rigler `r if (isTRUE(rig_own$own_area[rig_own$scenario == "sq"]) && !isTRUE(rig_own$own_area[rig_own$scenario == "a"])) "has its own attendance area today and none in either scenario" else "keeps an attendance area"` (`r SRC$scenarios`). Lent went through the same change in 2023, and it ended up with about half the enrollment PPS projected (527 projected, `r fmt_n(sc("lent", "enrolled_2025"))` in 2025). PPS's projection for Rigler doesn't show that risk. It has Rigler growing from `r fmt_n(gr("sq", "2027-28"))` students with no changes to `r fmt_n(gr("a", "2027-28"))` with Scott's program, and still `r fmt_n(gr("a", "2031-32"))` in 2031-32. That is Rigler's current enrollment plus the arriving program, as if its neighborhood families keep choosing it after new kindergartners from its area are assigned to Scott. The guide also projects Rigler's share of multilingual-learner students falling from `r fmt_pct(as.numeric(gsr("No changes", "pct_multilingual")))` to `r fmt_pct(as.numeric(gsr("Options A and B", "pct_multilingual")))`, and its share of students in poverty from `r fmt_pct(as.numeric(gsr("No changes", "pct_poverty")))` to `r fmt_pct(as.numeric(gsr("Options A and B", "pct_poverty")))`. The guides don't say whether Rigler-area families would keep any priority to enroll there.

The buildings have room. PPS's School Profiles dashboard gives each building's current functional capacity and its 2025-26 utilization. The 2021 Long-Range Facility Plan, which defines functional capacity as a building's classrooms times PPS's standard class size minus rooms set aside for other uses, is shown for comparison.

```{r}
#| label: tbl-capacity
sctx |>
  mutate(school = factor(school, c("Atkinson", "Creston", "Lent", "Rigler"))) |> arrange(school) |>
  select(school, functional_capacity_2021, functional_capacity, enrolled_2025, use_2025, projected_2027_a, use_2027_a) |>
  gt() |>
  cols_label(school = "", functional_capacity_2021 = "2021 plan", functional_capacity = "2025-26", enrolled_2025 = "Students", use_2025 = "Share of capacity",
             projected_2027_a = "Students", use_2027_a = "Share of capacity") |>
  tab_spanner("Functional capacity", c(functional_capacity_2021, functional_capacity)) |>
  tab_spanner("October 2025", c(enrolled_2025, use_2025)) |>
  tab_spanner("PPS projection, 2027-28 (Scenario A)", c(projected_2027_a, use_2027_a)) |>
  fmt_number(c(functional_capacity_2021, functional_capacity, enrolled_2025, projected_2027_a), decimals = 0) |> fmt_percent(c(use_2025, use_2027_a), decimals = 0) |>
  sub_missing(missing_text = "closes") |>
  tab_header(title = "Building capacity at the Southeast Spanish sites") |>
  tab_source_note("Shares use PPS's current (2025-26) capacity. PPS's space-use goal is 70–100% of available room (school guides). Creston's 2021 figure counted its annex. Enrollment is October 2025, kindergarten through grade 5, all programs; it matches the dashboard's enrollment for every school.") |>
  tab_source_note(md(cite("profiles25", "lrfp", "by_grade", "guides")))
```

Atkinson uses about `r fmt_pct(sc("atkinson", "use_2025"))` of its capacity, leaving room for roughly `r fmt_n(round(sc("atkinson", "functional_capacity") - sc("atkinson", "enrolled_2025"), -1))` more students, about `r fmt_n(round((sc("atkinson", "functional_capacity") - sc("atkinson", "enrolled_2025")) / 6))` per grade. That would fit a second Spanish class in every grade without moving anyone, though not a full doubling of today's program, which has `r fmt_n(sum(by_grade$enrollment[by_grade$key == "atkinson" & by_grade$year == "2025-26" & by_grade$language %in% "Spanish" & by_grade$grade %in% c("K", 1:5)]))` Spanish immersion students in K-5. (PPS's 2021 plan gave Atkinson more room: `r fmt_n(sc("atkinson", "functional_capacity_2021"))` seats.) A whole-school Spanish program would not fit unless Atkinson's `r fmt_n(sum(by_grade$enrollment[by_grade$key == "atkinson" & by_grade$year == "2025-26" & !(by_grade$language %in% "Spanish")]))` neighborhood-program students moved out, which is what went wrong at Lent. Lent is the opposite case: even after Atkinson's program moves in, PPS projects it at `r fmt_n(sc("lent", "projected_2027_a"))` students, `r fmt_pct(sc("lent", "use_2027_a"))` of its capacity, still short of PPS's own space-use goal; its guide marks the goal as not met.

That points to two other ways to serve Southeast families, both closer to where the demand is:

- **Add a second Spanish class at Atkinson**, filled from its waitlist, and keep its neighborhood program. This fits in the current building, as long as Creston's neighborhood children aren't also sent there.
- **Make Creston's building a Southeast Spanish immersion school.** Creston closes in both scenarios, and PPS already plans to send its neighborhood children to Atkinson (`r SRC$guides`). Moving Atkinson's immersion program into Creston instead of Lent would move it `r sprintf("%.1f", sc("creston", "mi_from_atkinson"))` miles rather than `r sprintf("%.1f", sc("lent", "mi_from_atkinson"))`, and the building has room for about `r fmt_n(round(sc("creston", "functional_capacity") / 6))` students per grade at PPS's current capacity of `r fmt_n(sc("creston", "functional_capacity"))` (`r SRC$profiles25`), more than twice Atkinson's Spanish program today. Creston would need work first: it is an unreinforced masonry building that PPS has targeted for seismic retrofit, according to the PPS facility data compiled in [meub/pps-data](https://github.com/meub/pps-data).

Either way, Lent could get back a neighborhood program serving its own families, many of them Spanish speakers. One question remains. Two-way immersion needs native speakers: about `r fmt_n(sc("creston", "spanish_517_within_1_5mi"))` school-age Spanish speakers live within 1.5 miles of Creston, fewer than near Atkinson or Lent, so a Creston site would need to recruit them.

PPS's case for whole-school sites is that teachers need same-grade partners. A second class per grade at Atkinson would give them one. See [Research](research.qmd#sec-single) for what the evidence says about whole-school and strand programs.

## 9. What a different solution could look like {#sec-alternatives}

The data points to options that would take less from immersion families. Each of these deserves to be evaluated before a decision:

- **Keep high-demand programs where the families are.** Atkinson and Beach are oversubscribed every year. Consolidate only where demand is low (James John, Sitton, Clark).
- **In the Southeast, grow Spanish immersion where the demand is**: a second class at Atkinson, or a Southeast immersion school in Creston's building, 0.8 miles away instead of 2.4 to Lent. **Restore Lent's neighborhood program** ([§8](#sec-southeast)).
- **Keep Rigler's neighborhood boundary**, the feature that makes it work today, or at least guarantee its neighborhood families a seat.
- **Move the neighborhood program instead of the immersion strand.** The plan already does this at César Chávez, where non-immersion students move to Rosa Parks so the school can become an immersion hub.
- **Prefer two-class strands to whole-school conversions** where the goal is teacher collaboration. No study we found shows whole-school programs outperform strands ([Research](research.qmd)).
- **Fill "underenrolled" immersion classrooms from the waitlists** instead of relocating them.
- **Phase changes in and protect current families.** Let enrolled students finish where they are, and guarantee transportation to any relocated program.
- **Do the equity analysis PPS hasn't published.** Report every change by immersion enrollment, home language, English-learner status, race and income. Include home-to-school distance before and after. Publish the share of families the projections assume will follow each moved program (about `r round(100 * mean(fol1$followed))`%, judging by the school guides), and track how many actually do. PPS has the student-level data to do this.

## Details {#sec-distance-detail}

### Estimated home-to-school distance

Children are placed by 2020 census block. For closures, children living in the closing school's attendance area are used, and their new school is the one assigned on the scenario map. For immersion programs, children are those nearest the program's current site of that language (an assumption), and their new school is the program's new site.

```{r}
#| label: fig-home
#| fig-height: 9
#| fig-cap: "Model estimate, not measured: average home-to-school distance before and after each change in Scenario A, for children living in the closing school's attendance area, or in the area nearest the immersion program's current school (an assumption about where immersion families live). Hover for details."
hd <- home |> filter(scenario == "a") |>
  mutate(grp = if_else(type == DLI, "Immersion program moves", "School closures"),
         label = if_else(type == DLI, paste0(from_school, " → ", to_school), paste0(from_school, " (closes)")),
         full = if_else(type == DLI, paste0(label, " (", language, ")"), label)) |>
  arrange(grp, change_mi)
hd$label <- factor(wrap_label(hd$label, 26), wrap_label(hd$label, 26))
col <- c("Immersion program moves" = PAL$light[["blue"]], "School closures" = PAL$muted)
p <- plot_ly(hd, y = ~label) |>
  add_segments(x = ~before_mi, xend = ~after_mi, yend = ~label, showlegend = FALSE, hoverinfo = "skip",
               line = list(color = PAL$grid, width = 2)) |>
  add_markers(x = ~before_mi, name = "Today", marker = list(color = "white", size = 9, line = list(color = PAL$ink2, width = 2)),
              text = ~sprintf("<b>%s</b><br>today: %.1f mi", full, before_mi), hoverinfo = "text")
for (k in names(col)) {
  p <- add_markers(p, data = filter(hd, grp == k), x = ~after_mi, name = paste("After:", tolower(k)),
                   marker = list(color = col[[k]], size = 10, line = list(color = "white", width = 2)),
                   text = ~sprintf("<b>%s</b><br>today %.1f mi → after %.1f mi (+%.1f)", full, before_mi, after_mi, change_mi), hoverinfo = "text")
}
viz_layout(p, xtitle = "Miles from home (estimate)") |>
  layout(xaxis = list(rangemode = "tozero"))
```

::: {.source}
`r cite("scenarios", "blocks")`
:::

```{r}
#| label: tbl-sensitivity
read_data("equity_home_sensitivity.csv") |>
  mutate(scenario = paste("Scenario", toupper(scenario))) |>
  select(scenario, variant, before, after, change) |>
  gt(groupname_col = "scenario") |>
  cols_label(variant = "Version of the estimate", before = "Today", after = "After", change = "Increase") |>
  fmt_number(c(before, after, change), decimals = 1) |>
  tab_header(title = "Estimated home-to-school distance, miles: how sensitive is it?", subtitle = "Model estimate; averages over affected students") |>
  tab_source_note("All versions assume immersion families live in the area nearest their program's current site of that language; that assumption cannot be tested without student addresses. \"Weighted by Spanish-speaking children\" places children by ACS school-age Spanish speakers rather than all children.") |>
  tab_source_note(md(cite("scenarios", "blocks", "acs", "by_grade")))
```

**Checked against where students live.** PPS's *School Enrollment by Neighborhood of Residence* report counts each program's students living in the school's own attendance area and, for the rest, the high-school area they live in. Those students are placed on census blocks within that area, either spread over all its children or over the half living closest to the program's current school.

```{r}
#| label: tbl-home-check
read_data("equity_home_check.csv") |> filter(scenario == "a") |>
  mutate(move = paste0(from_school, " → ", to_school, " (", language, ", ", band, ")"), own = own_area / students) |>
  arrange(desc(moving)) |>
  select(move, students, own, before_mi, change_mi, change_mi_nearer, model_before_mi, model_change_mi) |>
  gt() |>
  cols_label(move = "Move", students = "Students", own = "Live in school's own area", before_mi = "Today",
             change_mi = "Increase (area average)", change_mi_nearer = "Increase (nearer half)",
             model_before_mi = "Today", model_change_mi = "Increase") |>
  tab_spanner("From where students live (PPS report)", c(before_mi, change_mi, change_mi_nearer)) |>
  tab_spanner("Model", c(model_before_mi, model_change_mi)) |>
  fmt_number(students, decimals = 0) |> fmt_percent(own, decimals = 0) |>
  fmt_number(c(before_mi, change_mi, change_mi_nearer, model_before_mi, model_change_mi), decimals = 1) |>
  tab_header(title = "Estimated home-to-school distance: model vs where students live (Scenario A, miles)") |>
  tab_source_note("Students: all students in the sending program, October 2025 (César Chávez's row covers grades K-8). Students living outside their school's area are known only by high-school area; Jefferson's three sub-areas are approximated by which of the Grant, McDaniel and Roosevelt areas is nearest. Out-of-district and undetermined addresses are left out. Straight-line distances.") |>
  tab_source_note(md(cite("by_residence", "scenarios", "blocks")))
```

### Every closure and immersion move

```{r}
#| label: tbl-changes
chg |> filter(scenario == "a") |>
  mutate(type = if_else(type == DLI, paste(language, "immersion moves"), "Schools that close"),
         change = if_else(grepl("immersion", type), paste(from_school, "→", to_school), from_school)) |>
  arrange(type != "Schools that close", type, desc(pct_spanish_home)) |>
  select(type, change, band, students, pct_spanish_home, pct_non_english_home, school_miles) |>
  gt(groupname_col = "type") |>
  cols_label(change = "Change", band = "Grades", students = "Students", pct_spanish_home = "Spanish at home",
             pct_non_english_home = "Non-English at home", school_miles = "School moves (mi)") |>
  fmt_number(students, decimals = 0) |> fmt_percent(c(pct_spanish_home, pct_non_english_home), decimals = 0) |>
  fmt_number(school_miles, decimals = 1) |>
  tab_header(title = "Every closure and immersion move, Scenario A") |>
  tab_source_note("Students: October 2025 enrollment; for immersion moves, students in the moving language program in the grades that move; for closures, the school's non-immersion students (its immersion students are counted under the program move). Home language: share of those students. School moves: straight-line distance from the current school to the new one; for closures split among several receiving schools, weighted by how much of the closing school's attendance area each receives.") |>
  tab_source_note(md(cite("by_grade", "by_lang", "scenarios")))
```

### Caveats

- Lottery figures count applications, not children.
- Race and income are whole-school figures; PPS doesn't publish them by program.
- Home-to-school distances and families not following are estimates.
- Boundary-change moves are not included.
- This is an independent analysis, not a PPS publication.

See [Sources & methods](methods.qmd#equity) for full definitions.
