---
title: "Equity: who the changes fall on"
subtitle: "PPS reports similar move rates across student groups. Program moves and closures affect very different students."
---

```{r}
source("R/site_data.R")
source("R/charts.R")
library(gt)
s44 <- read_data("pps_slide44.csv")
grp <- read_data("equity_groups.csv")
chg <- read_data("equity_changes.csv")
home <- read_data("equity_home_distance.csv")
homes <- read_data("equity_home_distance_summary.csv")
g <- function(sc, ty, col) grp[[col]][grp$scenario == sc & grp$type == ty]
DLI <- "Immersion program moves"; CLO <- "School closures (non-immersion students)"; ALL <- "All PPS students"
h <- function(sc, ty, col) homes[[col]][homes$scenario == sc & homes$type == ty]
```

## What PPS reported

PPS's equity check for the scenarios is one slide: the share of kindergarten–8th grade students in each group who would change schools. Its message is that impacts are spread across groups. The numbers are close, but not equal: **multilingual learners are the most-moved group in both scenarios**, about four points above the district average.

```{r}
#| label: tbl-slide44
s44 |>
  gt() |>
  cols_label(group = "Student group", a_number = "Students", a_pct = "Share", b_number = "Students", b_pct = "Share") |>
  tab_spanner("Scenario A", c(a_number, a_pct)) |> tab_spanner("Scenario B", c(b_number, b_pct)) |>
  fmt_number(c(a_number, b_number), decimals = 0) |> fmt_percent(c(a_pct, b_pct), decimals = 0) |>
  tab_style(cell_text(weight = "bold"), cells_body(rows = group == "Multilingual Learners")) |>
  tab_source_note("As printed by PPS. The slide gives no year and does not define \"move\" (for example, whether a student whose immersion program relocates counts). Groups overlap.") |>
  tab_source_note(md(cite("slide44")))
```

The slide has no row for **Hispanic/Latino students**, **home language**, or **dual language immersion students**, and the board packet has no analysis of **how far** students would travel. The scenario memo acknowledges the trade-off without measuring it: "consolidating an underenrolled immersion program may lengthen some students' trips" (`r SRC$memo`).

District-wide move rates also blend very different kinds of change. Most moves come from boundary changes; within them sit two smaller groups that are directly displaced: students whose **school closes** and students whose **immersion program moves**. The rest of this page compares those two.

## Who is in the programs that move

```{r}
#| label: fig-who
#| fig-cap: "Share of students whose home language is Spanish, or any language other than English: students in immersion programs that move, students at schools that close (excluding their immersion programs), and all PPS students. October 2025 enrollment."
d <- grp |> mutate(type = factor(type, c(DLI, CLO, ALL), c("Immersion program moves", "School closures", "All PPS students")),
                   scen = paste0("Scenario ", toupper(scenario))) |>
  filter(!(type == "All PPS students" & scenario == "b")) |>
  mutate(scen = if_else(type == "All PPS students", "District", scen), label = paste0(type, if_else(scen == "District", "", paste0(", ", scen))))
d$label <- factor(d$label, rev(unique(d$label[order(d$type, d$scen)])))
tip <- function(x, what) sprintf("<b>%s</b><br>%s%% %s<br>%s students", d$label, round(100 * x), what, fmt_n(d$students))
p <- plot_ly(d, y = ~label) |>
  add_bars(x = ~pct_spanish_home, name = "Spanish at home", marker = list(color = PAL$light[["blue"]], line = list(color = "white", width = 2)),
           text = tip(d$pct_spanish_home, "speak Spanish at home"), hoverinfo = "text", textposition = "none") |>
  add_bars(x = ~pct_non_english_home, name = "Any non-English language at home", marker = list(color = PAL$light[["orange"]], line = list(color = "white", width = 2)),
           text = tip(d$pct_non_english_home, "speak a language other than English at home"), hoverinfo = "text", textposition = "none")
viz_layout(p, xtitle = "Share of students") |>
  layout(barmode = "group", bargap = 0.3, xaxis = list(tickformat = ".0%", rangemode = "tozero"), height = 380)
```

::: {.source}
`r cite("by_lang", "by_grade", "scenarios", "closures")`
:::

Students in the immersion programs that move are about **`r round(g("a", DLI, "pct_spanish_home") / g("a", CLO, "pct_spanish_home"))` times as likely** as students at closing schools to speak Spanish at home (`r fmt_pct(g("a", DLI, "pct_spanish_home"))` vs `r fmt_pct(g("a", CLO, "pct_spanish_home"))` in Scenario A), and about `r round(g("a", DLI, "pct_spanish_home") / g("a", ALL, "pct_spanish_home"))` times the district rate (`r fmt_pct(g("a", ALL, "pct_spanish_home"))`). Of the Spanish-speaking students directly displaced in Scenario A, by a closure or a program move, **`r fmt_pct(g("a", DLI, "spanish_home") / (g("a", DLI, "spanish_home") + g("a", CLO, "spanish_home")))` are in immersion programs** (`r fmt_n(g("a", DLI, "spanish_home"))` of `r fmt_n(g("a", DLI, "spanish_home") + g("a", CLO, "spanish_home"))`).

```{r}
#| label: tbl-changes
chg |> filter(scenario == "a") |>
  mutate(type = if_else(type == DLI, paste(language, "immersion moves"), "Schools that close"),
         change = if_else(grepl("immersion", type), paste(from_school, "→", to_school), from_school)) |>
  arrange(type != "Schools that close", type, desc(pct_spanish_home)) |>
  select(type, change, band, students, pct_spanish_home, pct_non_english_home, school_miles) |>
  gt(groupname_col = "type") |>
  cols_label(change = "Change", band = "Grades", students = "Students", pct_spanish_home = "Spanish at home",
             pct_non_english_home = "Non-English at home", school_miles = "School moves (mi)") |>
  fmt_number(students, decimals = 0) |> fmt_percent(c(pct_spanish_home, pct_non_english_home), decimals = 0) |>
  fmt_number(school_miles, decimals = 1) |>
  tab_header(title = "Every closure and immersion move, Scenario A") |>
  tab_source_note("Students: October 2025 enrollment; for immersion moves, students in the moving language program in the grades that move; for closures, the school's non-immersion students (its immersion students are counted under the program move). Home language: share of those students. School moves: straight-line distance from the current school to the new one; for closures split among several receiving schools, weighted by how much of the closing school's attendance area each receives.") |>
  tab_source_note(md(cite("by_grade", "by_lang", "scenarios")))
```

## How far families would travel

### Measured: how far the school moves

The firmer measure is the straight-line distance from each student's current school to the school they would attend, using the school locations on the PPS scenario maps. Students at closing schools mostly move to the next school over: on average `r sprintf("%.1f", g("a", CLO, "school_miles"))` miles in Scenario A. Immersion programs move farther: **`r sprintf("%.1f", g("a", DLI, "school_miles"))` miles** on average, with the Atkinson, Beach, James John and Sitton Spanish programs moving 1.6–2.5 miles (Scott → Rigler, at 0.7 miles, is the exception). Each change's distance is in the table above.

```{r}
#| label: tbl-school-miles
grp |> filter(type != ALL) |>
  mutate(type = if_else(type == DLI, "Immersion program moves", "School closures"), scenario = paste("Scenario", toupper(scenario))) |>
  select(type, scenario, school_miles) |>
  pivot_wider(names_from = scenario, values_from = school_miles) |>
  gt() |>
  cols_label(type = "") |>
  fmt_number(-type, decimals = 1) |>
  tab_header(title = "How far the school moves, miles (straight line)", subtitle = "Average over affected students") |>
  tab_source_note("Distance from the current school to the new school, averaged over each change's enrolled students (October 2025). For closures split among several receiving schools, weighted by the share of the closing school's attendance area each receives.") |>
  tab_source_note(md(cite("scenarios", "by_grade", "by_lang")))
```

### Estimated: change in distance from home

Distance from home is what families experience, but PPS doesn't publish where students live, so this part is a **model estimate**, not a measurement. Children are placed where the 2020 Census counted them (census blocks). For a closing school, that is its attendance area, which is a good guide to where its students live. For an immersion program it has to be **assumed**: each program's families are taken to live in the area nearer to it than to any other site of the same language. Real immersion families often come from farther away, and for a language with a single site (Vietnamese) that area is the whole district.

With that assumption, children near a closing school would be `r sprintf("%.1f", h("a", CLO, "change"))` miles farther from school on average (`r sprintf("%.1f", h("a", CLO, "before"))` → `r sprintf("%.1f", h("a", CLO, "after"))` miles), and children near a moving immersion program about `r sprintf("%.1f", h("a", DLI, "change"))` mile farther (`r sprintf("%.1f", h("a", DLI, "before"))` → `r sprintf("%.1f", h("a", DLI, "after"))` miles). The gap holds when the estimate is varied (below): `r sens <- read_data("equity_home_sensitivity.csv") |> filter(scenario == "a"); sprintf("+%.1f to +%.1f miles for immersion programs vs +%.1f to +%.1f for closures", min(sens$change[grepl("^Immersion", sens$variant)]), max(sens$change[grepl("^Immersion", sens$variant)]), min(sens$change[grepl("^Closures", sens$variant)]), max(sens$change[grepl("^Closures", sens$variant)]))`. Student address data from PPS would be needed to confirm it.

```{r}
#| label: fig-home
#| fig-height: 8
#| fig-cap: "Model estimate, not measured: average home-to-school distance before and after each change in Scenario A, for children living in the closing school's attendance area, or in the area nearest the immersion program's current school (an assumption about where immersion families live). Hover for details."
hd <- home |> filter(scenario == "a") |>
  mutate(grp = if_else(type == DLI, "Immersion program moves", "School closures"),
         label = if_else(type == DLI, paste0(from_school, " → ", to_school, " (", language, ")"), paste0(from_school, " (closes)"))) |>
  arrange(grp, change_mi)
hd$label <- factor(hd$label, hd$label)
col <- c("Immersion program moves" = PAL$light[["blue"]], "School closures" = PAL$muted)
p <- plot_ly(hd, y = ~label) |>
  add_segments(x = ~before_mi, xend = ~after_mi, yend = ~label, showlegend = FALSE, hoverinfo = "skip",
               line = list(color = PAL$grid, width = 2)) |>
  add_markers(x = ~before_mi, name = "Today", marker = list(color = "white", size = 9, line = list(color = PAL$ink2, width = 2)),
              text = ~sprintf("<b>%s</b><br>today: %.1f mi", label, before_mi), hoverinfo = "text")
for (k in names(col)) {
  p <- add_markers(p, data = filter(hd, grp == k), x = ~after_mi, name = paste("After:", tolower(k)),
                   marker = list(color = col[[k]], size = 10, line = list(color = "white", width = 2)),
                   text = ~sprintf("<b>%s</b><br>today %.1f mi → after %.1f mi (+%.1f)", label, before_mi, after_mi, change_mi), hoverinfo = "text")
}
viz_layout(p, xtitle = "Estimated average distance from home to school (miles, straight line)") |>
  layout(xaxis = list(rangemode = "tozero"))
```

::: {.source}
`r cite("scenarios", "blocks")`
:::

```{r}
#| label: tbl-sensitivity
read_data("equity_home_sensitivity.csv") |>
  mutate(scenario = paste("Scenario", toupper(scenario))) |>
  select(scenario, variant, before, after, change) |>
  gt(groupname_col = "scenario") |>
  cols_label(variant = "Version of the estimate", before = "Today", after = "After", change = "Increase") |>
  fmt_number(c(before, after, change), decimals = 1) |>
  tab_header(title = "Estimated home-to-school distance, miles: how sensitive is it?", subtitle = "Model estimate; averages over affected students") |>
  tab_source_note("All versions assume immersion families live in the area nearest their program's current site of that language; that assumption cannot be tested without student addresses. \"Weighted by Spanish-speaking children\" places children by ACS school-age Spanish speakers rather than all children.") |>
  tab_source_note(md(cite("scenarios", "blocks", "acs", "by_grade")))
```

## What this does and doesn't show

- **It shows** that the students PPS directly displaces through program moves are far more likely to be Spanish speakers than the students displaced by closures, and that their schools move farther (measured). That pattern is invisible in a district-wide move rate.
- **It does not show race or ethnicity.** PPS publishes each school's racial makeup but not its immersion program's, so the share of Hispanic/Latino students in moving programs can't be computed from public reports. Home language is the closest available measure.
- **Distances.** How far each school moves is measured from the scenario maps (straight line). How far families' trips grow is a model estimate: for immersion families it rests on an assumption about where they live, because PPS does not publish students' addresses. Both are straight-line, so real travel is longer.
- **What would settle it**: PPS has student-level data. A request for the number of students affected by each change by race/ethnicity, home language and immersion enrollment, and for average home-to-school distance before and after, would let PPS's slide 44 be checked directly.

See [Sources & methods](methods.qmd#equity) for details.
