---
title: "Lent and César Chávez"
subtitle: "Two Spanish immersion schools, from 2019 to PPS's 2031 projections"
---

```{r}
source("R/site_data.R")
source("R/charts.R")
library(gt)
he <- read_data("history_enrollment.csv")
hp <- read_data("history_projection.csv")
cr <- read_data("capture_rate.csv") |> filter(neighborhood %in% c("Lent", "César Chávez"), row_type == "school")
ebn <- read_data("enroll_by_neighborhood.csv") |> filter(school %in% c("Lent", "César Chávez"), program == "Spanish Immersion")
dem <- read_data("school_demographics.csv") |> filter(name %in% c("Lent", "César Chávez"), year == "2025-26")
prof <- read_data("school_profiles_2025.csv") |> mutate(key = school_key(school))
gs <- read_data("guide_school.csv") |> filter(school %in% c("Lent", "Cesar Chavez"))
fcl <- read_data("equity_forecast_check.csv")
v <- function(s, p, y) he$enrollment[he$school == s & he$program == p & he$year == y]
pj <- function(s, sc, y) hp$enrollment[hp$school == s & hp$scenario == sc & hp$year == y]
cap <- function(s, y, col = "own_neighborhood_school") cr[[col]][cr$neighborhood == s & cr$year == y]
capr <- function(s, y) cap(s, y) / cap(s, y, "total")
own <- function(s, y) { d <- ebn |> filter(school == s, year == y); sum(d$students[d$area_level == "own_neighborhood"]) / d$row_total[1] }
dm <- function(s, rt, col) dem[[col]][dem$name == s & dem$row_type == rt]
fc <- function(k) prof$functional_capacity[prof$key == k]
cc_g <- by_grade |> filter(key == "cesarchavez", year == "2025-26", grade != "PK") |>
  mutate(program = if_else(language %in% "Spanish", "Spanish immersion", "Neighborhood program"),
         band = if_else(grade %in% c("6", "7", "8"), "Grades 6-8", "Grades K-5"))
ccg <- function(p, b) sum(cc_g$enrollment[cc_g$program == p & cc_g$band == b])
```

Two schools show what the scenarios would mean for Spanish dual language immersion (DLI). In a DLI program, children learn in both English and Spanish. **Lent** is the precedent. In 2023, Portland Public Schools (PPS) moved Bridger's Spanish program into Lent. PPS also moved Lent's neighborhood program out, so Lent became an all-immersion school. **César Chávez** is the plan's main new hub, a school that takes in programs from other schools. In both scenarios, PPS would make it a regional Spanish immersion school for kindergarten through fifth grade (K-5). It would take in the immersion programs from Beach, Sitton and James John. Its neighborhood students would move to Rosa Parks, and its middle grades would move to George.

## 1. How enrollment has changed since 2019 {#sec-enrollment}

```{r}
#| label: fig-history
#| fig-height: 5.4
#| fig-cap: "October enrollment, kindergarten through grade 12, by program. Lent and Bridger were K-8 schools until 2021-22, when their middle grades moved out; César Chávez is still K-8. In 2023-24 Bridger's Spanish program moved to Lent and Lent's neighborhood program moved to Marysville. Hover for counts."
d <- he |> filter(program != "Total", !(school == "Bridger" & year >= "2023-24"), !(school == "Lent" & program == "Neighborhood" & year >= "2023-24")) |>
  mutate(series = paste(school, tolower(program)), series = sub("spanish", "Spanish immersion", series), series = sub("neighborhood$", "neighborhood program", series))
lev <- c("Lent Spanish immersion", "Bridger Spanish immersion", "Lent neighborhood program", "César Chávez Spanish immersion", "César Chávez neighborhood program")
col <- setNames(c(PAL$light[["blue"]], PAL$light[["aqua"]], PAL$light[["orange"]], PAL$light[["blue"]], PAL$light[["orange"]]), lev)
dash <- setNames(c("solid", "solid", "solid", "dot", "dot"), lev)
p <- plot_ly()
for (s in lev) {
  ds <- filter(d, series == s)
  p <- add_trace(p, data = ds, x = ~year, y = ~enrollment, type = "scatter", mode = "lines+markers", name = s,
                 line = list(color = col[[s]], width = 2, dash = dash[[s]]), marker = list(color = col[[s]], size = 8),
                 text = ~sprintf("<b>%s</b> %s<br>%s students (%s school)", series, year, fmt_n(enrollment), grades), hoverinfo = "text")
}
viz_layout(p, ytitle = "Students (K-12)") |>
  layout(xaxis = list(type = "category"), yaxis = list(rangemode = "tozero"),
         annotations = list(list(x = "2023-24", y = 360, text = "Bridger's program<br>moves to Lent", align = "left", showarrow = FALSE, xanchor = "left", xshift = 4, font = list(size = 11, color = PAL$ink2))))
```

::: {.source}
`r cite("psu_premove")`
:::

In 2022-23, before the move, the Spanish programs at Bridger and Lent together had `r fmt_n(v("Bridger", "Spanish", "2022-23") + v("Lent", "Spanish", "2022-23"))` students in grades K-5. The combined program at Lent opened with `r fmt_n(v("Lent", "Spanish", "2023-24"))` students and has `r fmt_n(v("Lent", "Spanish", "2025-26"))` today.

At César Chávez, the Spanish program has held steady at about `r round(mean(he$enrollment[he$school == "César Chávez" & he$program == "Spanish"]), -1)` students. Its neighborhood program has shrunk from `r fmt_n(v("César Chávez", "Neighborhood", "2019-20"))` to `r fmt_n(v("César Chávez", "Neighborhood", "2025-26"))` students. (The drop from 2020-21 to 2021-22 at Lent and Bridger came from their middle grades moving out. It does not mean those schools lost students.)

## 2. Lent after the 2023 move: fewer students than planned {#sec-lent}

**Enrollment fell short of every projection.** PPS's 2023 plan expected Lent to reach "an estimated 527 students in 2025–26" (`r SRC$audit_schools`, p. 34). Portland State University (PSU) also forecasts PPS enrollment. Its June 2023 forecast already included the move. It had Lent at `r fmt_n(fcl$forecast[fcl$target == "Lent" & fcl$vintage == "2023-24" & fcl$year == "2025-26"])` students that year (`r SRC$psu_premove`). Lent actually enrolled `r fmt_n(v("Lent", "Total", "2025-26"))`. Lent now enrolls about 31% fewer immersion kindergartners a year than the two schools averaged per grade before the move. Measured the same way, the district fell 14% ([details](scorecard.qmd#a-precedent-bridgers-spanish-program-moved-to-lent-in-2023)).

**Lent no longer serves its neighborhood.** A school's attendance area is the neighborhood it serves. In 2022-23, `r fmt_pct(capr("Lent", "2022-23"))` of PPS students living in Lent's area attended Lent. This share is called the capture rate. After the conversion, it fell to `r fmt_pct(capr("Lent", "2023-24"))`. Most of that drop was by design: PPS assigned Lent's non-immersion neighborhood children to Marysville. Today `r fmt_pct(own("Lent", "2025-26"))` of Lent's students live in its area. Lent's students are now `r fmt_pct(dm("Lent", "School", "pct_latino"))` Latino. By comparison, PPS students living in Lent's area are `r fmt_pct(dm("Lent", "Neighborhood", "pct_latino"))` Latino (`r SRC$demographics`).

**Staff and families describe a hard transition.** The equity audit records this account: "A lot of the parents who pushed from Bridger for their kids to come to the school didn't end up staying…they felt like it was out of their way [and that] their students didn't feel supported" (`r SRC$audit_schools`, p. 39). After 2022-23, 63% of Lent's staff left (p. 39). You can read more accounts on the [Equity page](equity.qmd#sec-optout).

```{r}
#| label: tbl-lent-capture
cr |> filter(neighborhood == "Lent") |>
  left_join(ebn |> filter(school == "Lent") |> group_by(year) |> summarise(own = sum(students[area_level == "own_neighborhood"]), students = first(row_total)), by = "year") |>
  transmute(year, total, own_neighborhood_school, capture = own_neighborhood_school / total, students, own_share = own / students) |>
  gt() |>
  cols_label(year = "Year", total = "PPS students living in Lent's area", own_neighborhood_school = "Attend Lent",
             capture = "Share (capture rate)", students = "Lent students", own_share = "Share living in Lent's area") |>
  tab_spanner("Lent's neighborhood", c(total, own_neighborhood_school, capture)) |>
  tab_spanner("Lent school", c(students, own_share)) |>
  fmt_number(c(total, own_neighborhood_school, students), decimals = 0) |> fmt_percent(c(capture, own_share), decimals = 0) |>
  sub_missing(missing_text = "–") |>
  tab_header(title = "Lent and its neighborhood, 2019-20 to 2025-26") |>
  tab_source_note("Capture rate: share of PPS students (K-12) living in the attendance area who attend the area's school. The 2023-24 drop is largely by design: Lent's non-immersion neighborhood children were assigned to Marysville, which PPS counts as another neighborhood school. Lent was K-8 until 2021-22. Where Lent's students live is published from 2023-24.") |>
  tab_source_note(md(cite("capture", "by_residence")))
```

## 3. César Chávez today: two programs in one school {#sec-chavez}

César Chávez is a K-8 school with a Spanish immersion program and a smaller neighborhood program. In October 2025, it had:

- `r fmt_n(ccg("Spanish immersion", "Grades K-5"))` immersion students in grades K-5
- `r fmt_n(ccg("Spanish immersion", "Grades 6-8"))` immersion students in grades 6-8
- `r fmt_n(ccg("Neighborhood program", "Grades K-5") + ccg("Neighborhood program", "Grades 6-8"))` students in its neighborhood program

`r fmt_pct(own("César Chávez", "2025-26"))` of its immersion students live in its own attendance area. Its capture rate is the share of PPS students living in its area who attend it. Now `r fmt_pct(capr("César Chávez", "2025-26"))` of PPS students living in its area attend it, up from `r fmt_pct(capr("César Chávez", "2019-20"))` in 2019-20.

```{r}
#| label: tbl-chavez
tibble(program = c("Spanish immersion", "Neighborhood program")) |>
  mutate(k5 = vapply(program, ccg, 0, b = "Grades K-5"), m = vapply(program, ccg, 0, b = "Grades 6-8"), total = k5 + m,
         fate = c("Stays; joined by Beach, Sitton and James John immersion (grades 6-8 move to George)", "Grades K-5 move to Rosa Parks; grades 6-8 to George")) |>
  gt() |>
  cols_label(program = "", k5 = "Grades K-5", m = "Grades 6-8", total = "Total", fate = "In Scenarios A and B") |>
  fmt_number(c(k5, m, total), decimals = 0) |>
  tab_header(title = "César Chávez, October 2025") |>
  tab_source_note(md(paste("Enrollment excludes pre-kindergarten. Scenario changes as described in César Chávez's school guide:", '"César Chávez would change from a K-8 to a focused K-5 regional Spanish DLI school."'))) |>
  tab_source_note(md(cite("by_grade", "guides")))
```

```{r}
#| label: tbl-demographics
dem |> mutate(row_type = if_else(row_type == "School", "Attending the school", "Living in its attendance area")) |>
  select(name, row_type, enrollment, pct_latino, pct_white, pct_direct_cert, pct_ell, pct_hist_underserved) |>
  gt(groupname_col = "name") |>
  cols_label(row_type = "PPS students", enrollment = "Students", pct_latino = "Latino", pct_white = "White",
             pct_direct_cert = "Low-income (direct certification)", pct_ell = "English learners", pct_hist_underserved = "Historically underserved") |>
  fmt_number(enrollment, decimals = 0) |> fmt_percent(starts_with("pct"), decimals = 0) |>
  tab_header(title = "Who attends, and who lives nearby, October 2025") |>
  tab_source_note("Whole-school figures, kindergarten through grade 12. \"Living in its attendance area\" counts PPS students who live in the school's area, whichever school they attend.") |>
  tab_source_note(md(cite("demographics")))
```

## 4. What the scenarios would do to both schools {#sec-scenarios}

```{r}
#| label: fig-projection
#| fig-height: 5.2
#| fig-cap: "Total enrollment: October actuals through 2025-26, then PPS's projections from its school guides, with no changes and under Scenarios A and B (identical for these two schools). Hover for counts."
act <- he |> filter(program == "Total", school %in% c("Lent", "César Chávez"), year >= "2021-22") |> transmute(school, year, enrollment, line = "Actual")
prj <- hp |> filter(scenario %in% c("sq", "a")) |> transmute(school, year, enrollment, line = if_else(scenario == "sq", "PPS projection, no changes", "PPS projection, Scenarios A and B"))
dd <- bind_rows(act, prj)
yrs <- c("2021-22", "2022-23", "2023-24", "2024-25", "2025-26", "2026-27", "2027-28", "2028-29", "2029-30", "2030-31", "2031-32")
lcol <- c(Lent = PAL$light[["blue"]], `César Chávez` = PAL$light[["orange"]])
ldash <- c(Actual = "solid", `PPS projection, no changes` = "dot", `PPS projection, Scenarios A and B` = "dash")
p <- plot_ly()
for (s in names(lcol)) for (l in names(ldash)) {
  ds <- filter(dd, school == s, line == l)
  if (!nrow(ds)) next
  p <- add_trace(p, data = ds, x = ~year, y = ~enrollment, type = "scatter", mode = "lines+markers", name = paste0(s, ": ", sub("^PPS projection, ", "PPS projection, ", l)),
                 line = list(color = lcol[[s]], width = 2, dash = ldash[[l]]), marker = list(color = lcol[[s]], size = 7),
                 text = ~sprintf("<b>%s</b> %s<br>%s: %s students", school, year, line, fmt_n(enrollment)), hoverinfo = "text")
}
viz_layout(p, ytitle = "Students") |> layout(xaxis = list(type = "category", categoryarray = yrs, categoryorder = "array"), yaxis = list(rangemode = "tozero"))
```

::: {.source}
`r cite("psu_premove", "guides")`
:::

PPS projects that **Lent** would grow to `r fmt_n(pj("Lent", "a", "2027-28"))` students in 2027-28 once Atkinson's program arrives. With no changes, PPS projects `r fmt_n(pj("Lent", "sq", "2027-28"))`. Lent's school guide says the move "brings Lent's student numbers above steady target levels (400+ students)" (`r SRC$guides`). Even so, Lent would fill only `r fmt_pct(pj("Lent", "a", "2027-28") / fc("lent"))` of its capacity of `r fmt_n(fc("lent"))` students (`r SRC$profiles25`). Capacity is the number of students PPS says the building can serve. PPS's goal is for schools to use 70–100% of their space.

**César Chávez** would have `r fmt_n(pj("César Chávez", "a", "2027-28"))` students, all in Spanish immersion, compared with `r fmt_n(pj("César Chávez", "sq", "2027-28"))` with no changes. That is `r fmt_pct(pj("César Chávez", "a", "2027-28") / fc("cesarchavez"))` of its capacity of `r fmt_n(fc("cesarchavez"))`. The guide projects that its share of students in poverty would fall from `r fmt_pct(as.numeric(gs$pct_poverty[gs$school == "Cesar Chavez" & gs$option == "No changes"]))` to `r fmt_pct(as.numeric(gs$pct_poverty[gs$school == "Cesar Chavez" & gs$option == "Options A and B"]))`. Its share of students from underserved racial groups would fall from `r fmt_pct(as.numeric(gs$pct_underserved_race[gs$school == "Cesar Chavez" & gs$option == "No changes"]))` to `r fmt_pct(as.numeric(gs$pct_underserved_race[gs$school == "Cesar Chavez" & gs$option == "Options A and B"]))`.

These projections depend on families following their programs. PPS assumes most will. In the moves where we can measure it, PPS's projections have about 80% of students arriving at the new school ([Equity §5](equity.qmd#sec-optout)). After the 2023 move, Lent ended up at about half of the 527 students PPS projected.

::: {.callout-note collapse="true"}
## Caveats
- Enrollment is PSU's count of students enrolled in October, kindergarten through grade 12. We use the count as first reported in each forecast edition. PPS's own totals for Lent include pre-kindergarten (308 in 2023-24).
- Lent and Bridger were K-8 schools until 2021-22, so earlier totals include middle grades.
- Capture rates and demographics cover the whole school or the whole area. PPS doesn't publish race or income for immersion programs alone.
- PPS's projections come from school guides dated October 8, 2026. PPS labels the guides as drafts. Capacities are PPS's current (2025-26) figures from its School Profiles dashboard.
:::
