---
title: "Demand & scorecard"
subtitle: "Where families apply for immersion, and what each scenario does to it"
---

```{r}
source("R/site_data.R")
source("R/charts.R")
library(gt)
k_enr <- function(key, yr) fmt_n(sum(by_grade$enrollment[by_grade$key == key & by_grade$year == yr & by_grade$grade == "K" & by_grade$language %in% LANGUAGES]))
k_off <- function(key, yr) fmt_n(sum(lottery_k$offered[lottery_k$key == key & lottery_k$year == yr]))

k5_moves <- moves |> filter(band == "K-5", kind == "program") |>
  distinct(language, from_school, from_key, to_school, to_key, .keep_all = FALSE) |>
  mutate(in_a = TRUE)
k5_moves$in_b <- paste(k5_moves$from_key, k5_moves$to_key) %in%
  with(filter(moves, scenario == "b"), paste(from_key, to_key))

demand <- k5_moves |>
  left_join(select(lottery_avg, key, from_apps = applications, from_wait = waitlisted, from_years = years), by = c(from_key = "key")) |>
  left_join(select(lottery_avg, key, to_apps = applications, to_wait = waitlisted, to_years = years), by = c(to_key = "key")) |>
  arrange(desc(from_apps))
```

## Demand: programs move away from where families apply {#demand-programs-move-away-from-where-families-apply}

Families who want a kindergarten immersion seat apply through the PPS school-choice lottery (the exception is the neighborhood families of all-immersion Lent and Rigler; see the note below). The lottery reports show how many families applied to each program, how many got an offer and how many were waitlisted. Averaged over the last five lotteries (`r min(LOTTERY_YEARS)` to `r max(LOTTERY_YEARS)`), the Spanish programs that would move include the two with the **most applications and longest waitlists in the district**, Atkinson and Beach, and they would move to schools that have drawn far fewer applicants.

```{r}
#| label: fig-demand
#| fig-height: 5.4
#| fig-cap: "Kindergarten lottery applications per year (5-year average) at each program that moves, and at the school it moves to. Hover for offers and waitlists. Lent's figure counts only applicants from outside its neighborhood (see note below)."
d <- demand |> filter(!is.na(from_apps)) |>
  mutate(label = paste0(from_school, " → ", to_school, " (", language, if_else(in_b, "", ", A only"), ")"),
         label = wrap_label(label), label = factor(label, levels = rev(label)))
hover <- function(school, apps, wait, role) sprintf("<b>%s</b> (%s)<br>%s applications/yr<br>%s waitlisted/yr", school, role, fmt_n(apps), fmt_n(wait))
p <- plot_ly(d, y = ~label) |>
  add_segments(x = ~from_apps, xend = ~coalesce(to_apps, from_apps), yend = ~label, showlegend = FALSE,
               line = list(color = PAL$grid, width = 2), hoverinfo = "skip") |>
  add_markers(x = ~from_apps, name = "School losing the program", marker = list(color = PAL$light[["blue"]], size = 11, line = list(color = "white", width = 2)),
              text = ~hover(from_school, from_apps, from_wait, "losing the program"), hoverinfo = "text") |>
  add_markers(data = filter(d, !is.na(to_apps)), x = ~to_apps, name = "School receiving it (its own lottery today)",
              marker = list(color = PAL$light[["orange"]], size = 11, line = list(color = "white", width = 2)),
              text = ~hover(to_school, to_apps, to_wait, "receiving the program"), hoverinfo = "text")
viz_layout(p, xtitle = "Applications per year", ytitle = NULL) |>
  layout(xaxis = list(rangemode = "tozero"), yaxis = list(tickfont = list(size = 12)))
```

::: {.source}
`r cite("lottery", "scenarios")`
:::

Schools with no orange dot (Rigler, Boise-Eliot/Humboldt, Vestal) have no immersion lottery today: Boise-Eliot/Humboldt and Vestal would be new sites, and Rigler fills its Spanish program without appearing in the lottery reports.

::: {.callout-warning}
## Lottery numbers understate demand at whole-school immersion schools
At **Lent** and **Rigler**, every student is in Spanish immersion, so children from the school's own neighborhood enroll without entering the lottery; only families from outside the neighborhood apply. In 2025-26 Lent enrolled `r k_enr("lent", "2025-26")` immersion kindergartners but made `r k_off("lent", "2025-26")` lottery offers, and Rigler enrolled `r k_enr("rigler", "2025-26")` with no lottery at all. So Lent's orange dot above counts only out-of-neighborhood applicants and is not directly comparable with the sending schools, which enroll about as many kindergartners as they offer lottery seats.

`r cite("by_grade", "lottery")`
:::

### A precedent: Bridger's Spanish program moved to Lent in 2023 {#a-precedent-bridgers-spanish-program-moved-to-lent-in-2023}

```{r}
#| label: tbl-bridger
kg <- read_data("class_size_kg.csv") |> mutate(key = school_key(school))
k_actual <- by_grade |> filter(grade == "K", language == "Spanish") |> group_by(year, key) |> summarise(k_actual = sum(enrollment), .groups = "drop")
bl <- profiles |> filter(key %in% c("bridgercreativescience", "lent")) |>
  select(year, key, school, imm_spanish) |>
  inner_join(select(kg, year, key, kg_avg_class_size), by = c("year", "key")) |>
  left_join(k_actual, by = c("year", "key")) |>
  mutate(per_grade = imm_spanish / 6, k_est = coalesce(k_actual, per_grade), classes = k_est / kg_avg_class_size)
bl_year <- bl |> group_by(year) |>
  summarise(schools = paste0(school, " ", imm_spanish, collapse = " + "), per_grade = sum(per_grade),
            k_actual = if (all(is.na(k_actual))) NA_real_ else sum(k_actual, na.rm = TRUE),
            kg_size = paste(kg_avg_class_size, collapse = " / "), classes = sum(classes), .groups = "drop")
before <- mean(bl_year$per_grade[bl_year$year < "2023-24"])
after_k <- mean(bl_year$k_actual, na.rm = TRUE)
# Same baseline on both sides: an average across grades K-5 before the move vs kindergarten after.
# District grades K-5, October 2022 (PPS Enrollment - Summary Comparison, Oct 3 2022).
district_k5_2022 <- c(K = 3152, `1` = 3376, `2` = 3319, `3` = 3502, `4` = 3485, `5` = 3358)
district_k_after <- by_grade |> filter(grade == "K", year %in% c("2024-25", "2025-26")) |> group_by(year) |> summarise(n = sum(enrollment)) |> pull(n)
district_chg <- mean(district_k_after) / mean(district_k5_2022) - 1
district_chg_kk <- mean(district_k_after) / district_k5_2022[["K"]] - 1
bl_year |>
  gt() |>
  cols_label(year = "Year", schools = "Spanish immersion students", per_grade = "Per grade (÷ 6)",
             k_actual = "Kindergarten (actual)", kg_size = "Avg K class size", classes = "Est. K immersion classes") |>
  fmt_number(c(per_grade, classes), decimals = 1) |> fmt_number(k_actual, decimals = 0) |>
  sub_missing(missing_text = "–") |>
  tab_source_note("Classes = kindergarten immersion students (actual where published, otherwise the per-grade average) ÷ average kindergarten class size at the school (all kindergarten classes, both programs). An estimate, not a count.") |>
  tab_source_note(md(cite("profiles", "by_grade", "class_size", "district_k")))
```

PPS has consolidated Spanish immersion before. Until 2022-23 it ran at both **Bridger** and **Lent**, each a K-5 school with an immersion and a neighborhood program. In 2023-24 Bridger's program moved to Lent, which became all-immersion (Bridger became the K-8 Bridger Creative Science). Before the move the two schools had about **`r fmt_n(before)` Spanish immersion students per grade**, roughly three kindergarten classes. Lent has since enrolled **`r fmt_n(after_k)` immersion kindergartners a year** on average (2024-25 and 2025-26), two classes: about `r fmt_pct(1 - after_k / before)` fewer. Measured the same way district-wide (the October 2022 average across grades K-5 against kindergarten in 2024-25 and 2025-26), enrollment fell about `r fmt_pct(-district_chg)`; kindergarten to kindergarten, the district fell `r fmt_pct(-district_chg_kk)`.

This is an estimate: before 2024-25 PPS publishes immersion counts by school but not by grade, so "per grade" is the school's immersion total divided by six grades. District-wide, kindergarten classes run smaller than grades 1–5, so an all-grades average likely overstates the old kindergarten size; that is why the district comparison above uses the same all-grades baseline. Lottery applications can't be used for this comparison, because Lent's neighborhood families stopped applying once it became all-immersion (see the note above).

### Spanish waitlists over time {#spanish-waitlists-over-time}

```{r}
#| label: fig-waitlist
#| fig-height: 5
#| fig-cap: "Kindergarten waitlist at each Spanish immersion program, by lottery year. Blue: program moves away in both scenarios. Orange: school receives a moved program. Gray: unchanged."
sending <- unique(demand$from_key[demand$language == "Spanish"])
receiving <- unique(demand$to_key[demand$language == "Spanish"])
w <- lottery_k |> filter(language == "Spanish", year %in% LOTTERY_YEARS) |>
  mutate(role = case_when(key %in% sending ~ "Program moves away", key %in% receiving ~ "Receives a moved program", TRUE ~ "Unchanged"),
         col = c("Program moves away" = PAL$light[["blue"]], "Receives a moved program" = PAL$light[["orange"]], "Unchanged" = PAL$muted)[role])
p <- plot_ly()
for (s in unique(w$key)) {
  ws <- filter(w, key == s)
  p <- add_trace(p, data = ws, x = ~year, y = ~waitlisted, type = "scatter", mode = "lines+markers",
                 name = ws$role[1], legendgroup = ws$role[1],
                 showlegend = !(ws$role[1] %in% w$role[match(unique(w$key), w$key)][seq_len(match(s, unique(w$key)) - 1)]),
                 line = list(color = ws$col[1], width = 2), marker = list(color = ws$col[1], size = 8),
                 text = ~sprintf("<b>%s</b> %s<br>%s applications, %s offered, %s waitlisted", school, year, applications, offered, waitlisted),
                 hoverinfo = "text")
}
last <- w |> filter(year == max(year), waitlisted >= 8)
p <- add_annotations(p, data = last, x = ~year, y = ~waitlisted, text = ~school, xanchor = "left", showarrow = FALSE,
                     xshift = 8, font = list(size = 12, color = PAL$ink2))
viz_layout(p, ytitle = "Waitlisted applications") |> layout(xaxis = list(type = "category"), yaxis = list(rangemode = "tozero"))
```

::: {.source}
`r cite("lottery", "scenarios")`
:::

### Who is waitlisted {#who-is-waitlisted}

The 2026-27 kindergarten lottery breaks each program's applicants down by where they live and their language background. At the programs with the longest waitlists, the families turned away are mostly **English-speaking families from the surrounding area**, the families a move puts farther away. Native Spanish speakers, who get priority, are almost always offered seats.

```{r}
#| label: tbl-who
who <- lottery |>
  filter(year == "2026-27", level == "ES", language == "Spanish", !is_total, residence %in% c("NH", "RG", "TR"),
         background %in% c("English", "Heritage", "Native")) |>
  mutate(residence = recode(residence, NH = "Neighborhood", RG = "Regional area", TR = "Elsewhere in PPS")) |>
  group_by(school = coalesce(pps_school, school), residence, background) |>
  summarise(applications = sum(applications), offered = sum(offered), waitlisted = sum(waitlisted), .groups = "drop")
top <- who |> group_by(school) |> summarise(w = sum(waitlisted)) |> slice_max(w, n = 4) |> pull(school)
who |> filter(school %in% top) |>
  mutate(cell = sprintf("%d of %d", waitlisted, applications)) |>
  select(school, residence, background, cell) |>
  pivot_wider(names_from = background, values_from = cell, values_fill = "–") |>
  mutate(school = factor(school, levels = top),
         residence = factor(residence, c("Neighborhood", "Regional area", "Elsewhere in PPS"))) |>
  arrange(school, residence) |>
  select(school, residence, any_of(c("English", "Heritage", "Native"))) |>
  gt(groupname_col = "school") |>
  cols_label(residence = "Lives in", English = "English speakers", Heritage = "Heritage speakers", Native = "Native speakers") |>
  tab_spanner("Waitlisted of applied, by language background", any_of(c("English", "Heritage", "Native"))) |>
  tab_header(title = "2026-27 kindergarten Spanish immersion lottery", subtitle = "The four programs with the most waitlisted applicants") |>
  tab_source_note("Neighborhood = lives in the host school's attendance area (NH); Regional area = lives in the set of schools forming the high-school Spanish immersion assignment area (RG); Elsewhere = any other PPS address (TR). Language background as recorded by PPS on the application. A child can apply to several programs, so counts are applications, not children.") |>
  tab_source_note(md(cite("lottery26"))) |>
  cols_align("right", any_of(c("English", "Heritage", "Native")))
```

## Scorecard: every program move {#scorecard-every-program-move}

```{r}
#| label: tbl-moves
sc <- moves |>
  group_by(kind, language, band, from_school, from_key, to_school, to_key, from_closes) |>
  summarise(scen = paste(toupper(sort(unique(scenario))), collapse = " & "), .groups = "drop") |>
  rowwise() |>
  mutate(students = strand_students(from_key, language, band),
         from_her = heritage_share(from_key, language), to_her = heritage_share(to_key, language)) |>
  ungroup() |>
  left_join(select(lottery_avg, key, from_apps = applications, from_wait = waitlisted), by = c(from_key = "key")) |>
  mutate(across(c(from_apps, from_wait), \(x) if_else(band == "K-5", x, NA_real_)),
         move = paste(from_school, "→", to_school),
         note = case_when(kind == "grades" ~ "School becomes K-5", from_closes ~ "Sending school closes", TRUE ~ "")) |>
  arrange(factor(language, LANGUAGES), band, desc(students))
sc |>
  select(language, move, band, scen, students, from_apps, from_wait, from_her, to_her, note) |>
  gt(groupname_col = "language") |>
  cols_label(move = "Move", band = "Grades", scen = "Scenario", students = "Students",
             from_apps = "Applied", from_wait = "Waitlisted", from_her = "Sending", to_her = "Receiving", note = "Note") |>
  tab_spanner("Home language matches", c(from_her, to_her)) |>
  cols_width(move ~ px(230), note ~ px(170)) |>
  fmt_number(c(students, from_apps, from_wait), decimals = 0) |>
  fmt_percent(c(from_her, to_her), decimals = 0) |>
  sub_missing(missing_text = "–") |>
  tab_spanner("K lottery, per year", c(from_apps, from_wait)) |>
  tab_source_note("Students: October 2025 enrollment in the language program, in the grades that move. Home language: share of the program's immersion students whose home language is the program language (Chinese for Mandarin). Lottery demand is shown for kindergarten entry only; middle-grade programs have no comparable lottery.") |>
  tab_source_note(md(cite("scenarios", "by_grade", "by_lang", "lottery"))) |>
  opt_row_striping()
```

## Sites by language and grade band {#sites-by-language-and-grade-band}

```{r}
#| label: tbl-sites
sites |> filter(change != "removed" | is.na(change)) |>
  count(language, band, scenario) |>
  pivot_wider(names_from = scenario, values_from = n, values_fill = 0) |>
  select(language, band, sq, a, b) |>
  mutate(language = factor(language, LANGUAGES), band = factor(band, BANDS)) |> arrange(language, band) |>
  gt(groupname_col = "language") |>
  cols_label(band = "Grades", sq = "Status quo", a = "Scenario A", b = "Scenario B") |>
  tab_style(style = cell_text(weight = "bold"), locations = cells_body(columns = c(a, b), rows = a != sq | b != sq)) |>
  tab_source_note("A site is a school whose label on the scenario map names an immersion program in that language and grade band. Bold: fewer or more sites than today.") |>
  tab_source_note(md(cite("scenarios")))
```

## Distance to the nearest site {#distance-to-the-nearest-site}

```{r}
#| label: tbl-access
read_data("access_summary.csv") |>
  select(group, scenario, within_1mi, mean_mi) |>
  pivot_wider(names_from = scenario, values_from = c(within_1mi, mean_mi)) |>
  rename_with(\(x) sub("within_1mi", "within1", x)) |>
  select(group, within1_sq, within1_a, within1_b, mean_mi_sq, mean_mi_a, mean_mi_b) |>
  mutate(group = factor(group, c("Spanish", "Spanish (ages 5-17)", "Mandarin", "Vietnamese", "Russian"))) |> arrange(group) |>
  gt() |>
  cols_label(group = "Home language", within1_sq = "Status quo", within1_a = "A", within1_b = "B",
             mean_mi_sq = "Status quo", mean_mi_a = "A", mean_mi_b = "B") |>
  tab_spanner("Within 1 mile of a K-5 site", starts_with("within1")) |>
  tab_spanner("Average distance (miles)", starts_with("mean_mi")) |>
  fmt_percent(starts_with("within1"), decimals = 0) |> fmt_number(starts_with("mean_mi"), decimals = 1) |>
  tab_source_note("People age 5+ who speak each language at home (ACS 2020-2024 tract estimates, spread across each tract's 2020 census blocks by population; school-age Spanish by population under 18); straight-line distance from each block to the nearest K-5 site of that language. For school-age Spanish speakers, allowing for Census sampling error, the 90% range is 44-57% within 1 mile today and 29-41% in Scenarios A and B. Mandarin counts all Chinese speakers; Russian counts Russian, Polish and other Slavic languages. Japanese has no tract-level data.") |>
  tab_source_note(md(cite("acs", "blocks", "scenarios")))
```
